---
title: "Appendix B"
author: "Kat Wilson"
date: "7/28/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r cars, include=FALSE}
#install.packages("MatchIt")
library(MatchIt)
library(broom)
library(tidyverse)

rm(list = ls()) #clear the environment
load("merged_participation_analysis.Rdata")
data <- merged_participation
table(data$Participated)

######
# PART 2- Change factor and data clearn
########

#For our data, we will want to choose the max years that the teachers participated
analysis_data <- aggregate(`Years of Service` ~ `Deidentified ID`, data, max)
analysis_data <- merge(analysis_data, merged_participation)
analysis_data <- unique(analysis_data, by = 'Deidentified ID')
table(analysis_data$Participated) #312 participated in tip
#change outcome to factor
analysis_data$Retained <- as.factor(analysis_data$Retained)

## redo the logistic regression models
model <- glm(Retained ~ Participated, data = analysis_data, family = binomial)
summary(model) #participated is a significant predictors of retention
model2 <- glm(Retained ~ Participated + `Employee Age`, data = analysis_data, family = binomial)
summary(model2)
model3 <- glm(Retained ~ Participated + Strata, data = analysis_data, family = binomial)
summary(model3)
model4 <- glm(Retained ~ Participated + Ethnicity, data = analysis_data, family = binomial)
summary(model4) ##Interactions of Ethnicity

######
# PART 3- Visualize Distributions/Matchings
########

#numerical variables
data <- analysis_data %>%
  group_by(Participated) %>%
  summarise(n_parts = n(),
            mean_years_service = mean(`Years of Service`),
            mean_age = mean(`Employee Age`))
data
table(analysis_data$Strata, analysis_data$Participated)
table(analysis_data$`Employee Age`, analysis_data$Participated)
#categorical variables
table(analysis_data$Title, analysis_data$Participated)
table(analysis_data$Gender, analysis_data$Participated)
table(analysis_data$Ethnicity, analysis_data$Participated)
table(analysis_data$`HOME ORG CODE`, analysis_data$Participated)


######
# PART 4- Create binary indicators
########

analysis_data <- analysis_data %>%
  select(Ethnicity, Title, Gender, `Employee Age`, Participated,
         Retained, `HOME ORG CODE`, `Years of Service`, Strata,
         `Deidentified ID`)
#recode teachers
analysis_data$title_status <- ifelse(analysis_data$Title ==
                                       "TEACHER,FULL TIME"| analysis_data$Title ==
                                       "TEACHER,SPEC EDUCATION",
                                     
                                     'teacher','non_classroom_teacher')

#recode ethnicity
analysis_data$Ethnicity <- ifelse(analysis_data$Ethnicity ==
                                    "NAT AM/INUIT"| analysis_data$Ethnicity ==
                                    "OTHER" | analysis_data$Ethnicity ==
                                    "PREF NO DISC",
                                  
                                  'other',analysis_data$Ethnicity)

##recode employee age
analysis_data <- mutate(analysis_data,
                        Age = case_when(
                          `Employee Age` >= 70 ~ "70s",
                          `Employee Age` >=60 & `Employee Age`< 70 ~ "60s",
                          `Employee Age` >=50 & `Employee Age`< 60 ~ "50s",
                          `Employee Age` >=40 & `Employee Age`< 50 ~ "40s",
                          `Employee Age` >=30 & `Employee Age`< 40 ~ "30s",
                          `Employee Age` >=20 & `Employee Age`< 30 ~ "20s"
                        ))
table(analysis_data$Age, analysis_data$Participated)
table(analysis_data$Ethnicity, analysis_data$Participated)
table(analysis_data$title_status, analysis_data$Participated)

##add indicator variable for the categorical variables
analysis_data <- mutate(analysis_data,
                        title_binary= case_when(
                          title_status == "teacher" ~ 1,
                          title_status != "teacher" ~ 0
                        ))
analysis_data <- mutate(analysis_data,
                        strata_binary_10_15= case_when(
                          Strata == "10 to 15 years" ~ 1,
                          Strata != "10 to 15" ~ 0
                        ))
analysis_data <- mutate(analysis_data,
                        strata_binary_5_10= case_when(
                          Strata == "5 to less than 10 years" ~ 1,
                          Strata != "5 to less than 10 years" ~ 0
                        ))
analysis_data <- mutate(analysis_data,
                        strata_binary_less_5= case_when(
                          Strata == "less than 5" ~ 1,
                          Strata != "less than 5" ~ 0
                        ))
analysis_data <- mutate(analysis_data,
                        strata_binary_greater_15= case_when(
                          Strata == "greater than or equal to 15" ~ 1,
                          Strata != "greater than or equal to 15" ~ 0
                        ))
analysis_data <- mutate(analysis_data,
                        ethnicity_binary_african_am= case_when(
                          Ethnicity == "AFRICAN AM" ~ 1,
                          Ethnicity != "AFRICAN AM" ~ 0
                        ))
analysis_data <- mutate(analysis_data,
                        ethnicity_binary_caucasian =case_when(
                          Ethnicity == "CAUCASIAN" ~ 1,
                          Ethnicity != "CAUCASIAN" ~ 0
                        ))
analysis_data <- mutate(analysis_data,
                        ethnicity_binary_asian =case_when(
                          Ethnicity == "ASIAN" ~ 1,
                          Ethnicity != "ASIAN" ~ 0
                        ))
analysis_data <- mutate(analysis_data,
                        ethnicity_binary_latina =case_when(
                          Ethnicity == "LATINA/O" ~ 1,
                          Ethnicity != "LATINA/O" ~ 0
                        ))
analysis_data <- mutate(analysis_data,
                        ethnicity_binary_other =case_when(
                          Ethnicity == "other" ~ 1,
                          Ethnicity != "other" ~ 0
                        ))


analysis_data <- mutate(analysis_data,
                        gender_binary_female =case_when(
                          Gender == "F" ~ 1,
                          Gender != "F" ~ 0
                        ))
analysis_data <- mutate(analysis_data,
                        gender_binary_male =case_when(
                          Gender == "M" ~ 1,
                          Gender != "M" ~ 0
                        ))


######
# PART 5- Visualize Region of Common Support
########

###visualizations
library(ggplot2)
p <- ggplot(analysis_data, aes(fill = factor(Participated))) +
  geom_bar(position = "dodge") +
  scale_fill_discrete("Smoke")
p + aes(x=Ethnicity)
p + aes(x = Gender)
p + aes(x= title_status)

###look at distribution
ggplot(analysis_data, aes(x = `Employee Age`, fill = factor(Participated))) +
  geom_histogram(position = "identity") +
  scale_fill_discrete("Participated")

###or we can look at the quantiles within each group
tapply(analysis_data$`Employee Age`, analysis_data$Participated, 
       quantile, probs = seq(0.2,1,0.1))
tapply(analysis_data$`Years of Service`, analysis_data$Participated, 
       quantile, probs = seq(0.2,1,0.1))

##region of common support
analysis_data %>%
  ggplot(aes(x = `Employee Age`)) +
  geom_histogram(position = "identity") +
  facet_wrap(~Participated)
analysis_data %>%
  ggplot(aes(x = `Years of Service`)) +
  geom_histogram(position = "identity") +
  facet_wrap(~Participated)


```


## Matching Model 1

You can also embed plots, for example:

```{r pressure, echo=TRUE}

### Matching 1- All covariates in the matching model
##set seed
set.seed(1731)
class(analysis_data$Strata)
analysis_data$Gender <- as.factor(analysis_data$Gender)
analysis_data$employeeage <- as.numeric(analysis_data$`Employee Age`)
analysis_data$title_status <- as.factor(analysis_data$title_status)
analysis_data$Ethnicity <- as.factor(analysis_data$Ethnicity)
analysis_data$HomeOrgCode <- as.factor(analysis_data$`HOME ORG CODE`)
analysis_data$YearsOfService <- as.numeric(analysis_data$`Years of Service`)
nearest <- matchit(Participated ~
                     Gender+ Ethnicity + Strata + employeeage+ title_status
                     ,
                   family = "binomial",
                   method = "nearest",
                   caliper = 0.25,
                   data = analysis_data)
summary(nearest)
###plotting these
#plot(nearest)
nearest_matched <- match.data(nearest)
#now perform the statistical analysis
nearest_matched$Participated <- as.factor(nearest_matched$Participated)
model <- glm(Retained ~ Participated, data = nearest_matched,
             family = binomial)
summary(model)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


## Trimming

Trim the non TIP sample so that it is only tip in terms of Years of Experience.
There are no 32 years olds, 38 year olds, or 40 or older TIP teachers. However, these
grouped ARE represented in the non-TIP sample.


```{r pressure2, echo=TRUE}
set.seed(1731)
# trimming the sample
analysis_data_trimmed <- analysis_data %>%
  filter(YearsOfService %in% c(0:31, 33,34,35,36,37,39))

nearest <- matchit(Participated ~
                     Gender+ Ethnicity + Strata + employeeage+ title_status
                     ,
                   family = "binomial",
                   method = "nearest",
                   caliper = 0.25,
                   data = analysis_data_trimmed)
summary(nearest)
###plotting these
plot(nearest)
nearest_matched <- match.data(nearest)
#now perform the statistical analysis
nearest_matched$Participated <- as.factor(nearest_matched$Participated)
model <- glm(Retained ~ Participated, data = nearest_matched,
             family = binomial)
summary(model)
```




## Coursen the Ethnicity Variable

Coursen the Ethnicity variable. Match it by Caucasian and non-Caucasian


```{r pressure3, echo=TRUE}
analysis_data <- analysis_data %>%
  mutate(Caucasian_binary = ifelse(Ethnicity == "CAUCASIAN", 1, 0))
analysis_data$Caucasian_binary <- as.factor(analysis_data$Caucasian_binary)
set.seed(111731)
nearest <- matchit(Participated ~
                     Gender+ Caucasian_binary + Strata + employeeage+ title_status
                     ,
                   family = "binomial",
                   method = "nearest",
                   caliper = 0.25,
                   data = analysis_data)
summary(nearest)
###plotting these
#plot(nearest)
nearest_matched <- match.data(nearest)
#now perform the statistical analysis
nearest_matched$Participated <- as.factor(nearest_matched$Participated)
model <- glm(Retained ~ Participated, data = nearest_matched,
             family = binomial)
summary(model)

table(analysis_data$Age, analysis_data$Caucasian_binary)

```




## Coursen the Ethnicity Variable and use with the trimmed data

Coursen the Ethnicity variable. Match it by Caucasian and non-Caucasian


```{r pressure4, echo=TRUE}
analysis_data_trimmed <- analysis_data_trimmed %>%
  mutate(Caucasian_binary = ifelse(Ethnicity == "CAUCASIAN", 1, 0))
analysis_data_trimmed$Caucasian_binary <- as.factor(analysis_data_trimmed$Caucasian_binary)
set.seed(1731)
nearest <- matchit(Participated ~
                     Gender+ Caucasian_binary + Strata + employeeage+ title_status
                     ,
                   family = "binomial",
                   method = "nearest",
                   caliper = 0.25,
                   data = analysis_data_trimmed)
summary(nearest)
###plotting these
#plot(nearest)
nearest_matched <- match.data(nearest)
#now perform the statistical analysis
nearest_matched$Participated <- as.factor(nearest_matched$Participated)
model <- glm(Retained ~ Participated, data = nearest_matched,
             family = binomial)
summary(model)

```



## Assocation of Ethnicity and Strata


```{r pressure5, echo=TRUE}

ethnicity_strata <- analysis_data %>%
  group_by(Ethnicity, Strata) %>%
  summarise(n =n())
ethnicity_strata_spread <- spread(ethnicity_strata, key = "Strata", value = "n")
class(ethnicity_strata_spread$`less than 5`)
ethnicity_strata_spread <- ethnicity_strata_spread %>%
  mutate(total = `less than 5` + `5 to less than 10 years`+
           `10 to 15 years` +`greater than or equal to 15`) %>%
  mutate(percent_less_than5= `less than 5`/total,
         percent_5_to_10 = `5 to less than 10 years`/total,
         percent_10_15 = `10 to 15 years`/total,
         percent_greater_15 = `greater than or equal to 15`/total)
ethnicity_strata_spread <- ethnicity_strata_spread %>%
  select(Ethnicity, percent_less_than5, percent_5_to_10,
         percent_10_15, percent_greater_15)
library(scales)
ethnicity_strata_spread$percent_less_than5 <- percent(ethnicity_strata_spread$percent_less_than5)
ethnicity_strata_spread$percent_5_to_10 <- percent(ethnicity_strata_spread$percent_5_to_10)
ethnicity_strata_spread$percent_10_15 <- percent(ethnicity_strata_spread$percent_10_15)
ethnicity_strata_spread$percent_greater_15 <- percent(ethnicity_strata_spread$percent_greater_15)
ethnicity_strata_spread


ethnicity_strata <- analysis_data %>%
  group_by(Ethnicity, Strata)
ggplot(ethnicity_strata, aes(x = Ethnicity, y = YearsOfService)) +
  geom_boxplot() +
  facet_wrap(~Strata)
```



## Assocation of Ethnicity and Age


```{r pressure6, echo=TRUE}

ethnicity_age <- analysis_data %>%
  group_by(Ethnicity, Age) %>%
  summarise(n =n())

ethnicity_age_spread <- spread(ethnicity_age, key = "Age", value = "n")
class(ethnicity_age_spread$`less than 5`)
ethnicity_age_spread <- ethnicity_age_spread %>%
  mutate(total = `20s` + `30s`+
           `40s` +`50s` +`60s`+`70s`) %>%
  mutate(percent_20s= `20s`/total,
         percent_30s = `30s`/total,
         percent_40s = `40s`/total,
         percent_50s = `50s`/total, 
         percent_60s = `60s`/total,
         percent_70s = `70s`/total)
ethnicity_strata_spread <- ethnicity_age_spread %>%
  select(percent_20s, percent_30s, percent_40s, percent_50s,
         percent_60s, percent_70s)
library(scales)
ethnicity_age_spread$percent_20s <- percent(ethnicity_age_spread$percent_20s)
ethnicity_age_spread$percent_30s <- percent(ethnicity_age_spread$percent_30s)
ethnicity_age_spread$percent_40s <- percent(ethnicity_age_spread$percent_40s)
ethnicity_age_spread$percent_50s <- percent(ethnicity_age_spread$percent_50s)
ethnicity_age_spread$percent_60s <- percent(ethnicity_age_spread$percent_60s)
ethnicity_age_spread$percent_70s <- percent(ethnicity_age_spread$percent_70s)
ethnicity_age_spread

```



## Assocation of nonWhite and Age


```{r pressure7, echo=TRUE}

caucasian_binary_age <- analysis_data %>%
  group_by(Caucasian_binary, Age) %>%
  summarise(n =n())

caucasian_binary_age_spread <- spread(caucasian_binary_age, key = "Age", value = "n")
caucasian_binary_age_spread

```
