---
title: "Appendix A"
output:
  pdf_document: null
  html_document:
    df_print: paged
  word_document: default
  fig_caption: yes
---

## Overview

Upon cleaning and merging the TIP data, the final merged TIP data file is used to find summary statistics and compare percentages of completion between TIP participants and non TIP participants. The merged file is brought in from the script, "TIP merge and data clean" which produces one final data set for analysis. For more information on the nature of missing data and the merge, consult the prior script.

The merged TIP data file we use for the below figures and analyses has complete variables for 16,012 teachers, 312 of which participate in TIP, and 15700 who did not participate in TIP. Information on these teachers is collected across an 8 year period, from 2010 through 2018.

## Visualization

In order to evaluate the effectiveness of TIP participation on retention, we first look to get a picture of the average length of time spent in the district by teachers, as well as how retention patterns regularly play out in the Philadelphia school district. 

How long do TIP and non TIP participants stay in teaching profession, on average? Figure 1 breaks down these differences. TIP participation rates are relatively evenly split among four strata of teacher groups. These strata divide teachers into one of four groups, based on how long each teacher has served in the district. 

A majority of teachers have less than 5 years of service in the district. Around 30% of both TIP and non-TIP teachers have been in the school district for 5-15 years. 

Figure 2 breaks down the liklihood of retention for teachers based on their strata. The table sorts teachers by strata to compare how groups of teacher strata differ in retention rates. 

For Strata 1 (less than 5 years), the likelihood of retention is very similar for TIP and non TIP teachers, at around 64%. However, as we move into stratas 2 and 3, then the liklihood of retention increases for TIP teachers. This trend also holds for the final strata, where TIP teachers are more likely to stay in the profession than non-TIP teachers. 


```{r chunk1, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE}
library(dplyr)
library(kableExtra)
library(tidyverse)

#initial missingness changes
rm(list = ls()) #clear the environment

##remove the year variables out of each
load("data/merged_participation.Rdata")

#create a new table to work with that has one entry for each De ID

table_part <- merged_participation %>%
  filter(Participated == 1) %>%
  select(`Deidentified ID`, Title, Gender, Ethnicity, `Employee Age`, Organization, `Years of Service`, Retained, Strata, Participated) %>%
    group_by(`Deidentified ID`)
df.agg <- aggregate(`Employee Age` ~ `Deidentified ID`, table_part, max)
table_part <- merge(df.agg, table_part)
save(table_part, file = "data/table_part.Rdata")
table_nonpart <- merged_participation %>%
  filter(Participated == 0) %>%
  select(`Deidentified ID`, Title, Gender, Ethnicity, `Employee Age`, Organization, `Years of Service`, Retained, Strata, Participated)%>%
    group_by(`Deidentified ID`)
df.agg <- aggregate(`Employee Age` ~ `Deidentified ID`, table_nonpart, max)
table_nonpart <- merge(df.agg, table_nonpart)
save(table_nonpart, file = "data/table_nonpart.Rdata")

#create proportion tables
table_part_strat <- table_part %>%
  group_by(Strata) %>%
  dplyr::summarise(total = n())
#change to numeric
table_part_strat$total <- as.numeric(table_part_strat$total)
#add a new column with total
table_part_strat <- table_part_strat %>%
  mutate(overall = colSums(table_part_strat[,2])) %>%
  mutate(proportion = total/overall) %>%
  select(Strata, total, proportion) %>%
  mutate(order = NA)


#recode the factors
table_part_strat$strata <- as.factor(table_part_strat$Strata)
table_part_strat$strata <- recode_factor(table_part_strat$Strata, `1` = "less than 5", `2` = "5 to less than 10 years", `3` = "10 to 15 years", `4` = "greater than or equal to 15")
table_part_strat <- table_part_strat %>%
  arrange(Strata) %>%
  select(Strata, total, proportion)
library(scales)
table_part_strat$proportion <- percent(table_part_strat$proportion)

#print a kable table
#kable(table_part_strat, "latex", booktabs = T, caption = "Strata Table of Participants") %>%
 # kable_styling(latex_options = c("repeat_header", "striped", "hover"), font_size = 12, repeat_header_text = "TABLE")

#check length
#length(unique(table_part$`Deidentified ID`))

#same for non-participants
table_nonpart_strat <- table_nonpart %>%
  group_by(Strata) %>%
  dplyr::summarise(total = n())
#change to numeric
table_nonpart_strat$total <- as.numeric(table_nonpart_strat$total)
#add a new column with total
table_nonpart_strat <- table_nonpart_strat %>%
  mutate(overall = colSums(table_nonpart_strat[,2])) %>%
  mutate(proportion = total/overall) %>%
  select(Strata, total, proportion) %>%
  mutate(order = NA)
#recode the factors
table_nonpart_strat$Strata <- as.factor(table_nonpart_strat$Strata)
#levels(table_nonpart_strat$strata)
table_nonpart_strat$Strata <- recode_factor(table_nonpart_strat$Strata, `1` = "less than 5", `2` = "5 to less than 10 years", `3` = "10 to 15 years", `4` = "greater than or equal to 15")
table_nonpart_strat <- table_nonpart_strat %>%
  arrange(Strata) %>%
  select(Strata, total, proportion)
library(scales)
table_nonpart_strat$proportion <- percent(table_nonpart_strat$proportion)
table_part_strat_2<- table_part_strat %>%
  rename("Strata of TIP Participants" = `Strata`)
table_nonpart_strat_2 <- table_nonpart_strat %>%
  rename("Strata of non Participants" = `Strata`)
# combined <- bind_cols(table_part_strat_2, table_nonpart_strat_2) %>%
#   rename("total" = `total1`,
#          "proportion" = `proportion1`)

#print kable table
#kable(combined, "latex", booktabs = T, caption = "Strata Comparisons for TIP and non TIP participants") %>%
#  kable_styling(latex_options = c("repeat_header", "striped", "scale_down"), font_size = 12, repeat_header_text = "TABLE")




```



```{r chunk2, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE}
#create proportion tables
total_table <- bind_rows(table_nonpart, table_part)
table_proportions <- total_table %>%
  group_by(Strata, Retained, Participated) %>%
  dplyr::summarise(total = n()) %>%
  spread(key = Retained, value = total) %>%
  mutate(totals = `0` + `1`) %>%
  mutate(liklihood_not_retained = `0` / totals) %>%
  mutate(liklihood_retained = `1` / totals)

##make this proportion table
#recode the factors
total_table$strata <- as.factor(total_table$Strata)
#levels(total_table$strata)
total_table$strata <- recode_factor(total_table$Strata, `1` = "less than 5", `2` = "5 to less than 10 years", `3` = "10 to 15 years", `4` = "greater than or equal to 15")
table_proportions <- total_table %>%
  group_by(Strata, Retained, Participated) %>%
  dplyr::summarise(total = n()) %>%
  spread(key = Retained, value = total) %>%
  mutate(totals = `0` + `1`) %>%
  mutate(`Likelihood of leaving profession` = `0` / totals) %>%
  mutate(`Likelihood of retention` = `1` / totals) %>%
  arrange(Strata)
library(scales)
table_proportions$`Likelihood of leaving profession` <- percent(table_proportions$`Likelihood of leaving profession`)
table_proportions$`Likelihood of retention` <- percent(table_proportions$`Likelihood of retention`)
table_proportions$totals <- NULL
# #print kable table
# kable(table_proportions, "latex", booktabs = T, caption = "Liklihood of Retention") %>%
#   kable_styling(latex_options = c("repeat_header", "striped", "scale_down"), font_size = 12, repeat_header_text = "TABLE") %>%
#   column_spec(5:6, width = "4cm")


```

```{r chunk3.8, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE}


#append together
 table_plot <- bind_rows(table_nonpart, table_part)


#create proportion tables
table_proportions <- table_plot %>%
  filter(Participated == 1) %>%
  group_by(Strata) %>%
  dplyr::summarise(total = n())
#change to numeric
table_proportions$total <- as.numeric(table_proportions$total)
#add a new column with total
table_proportions <- table_proportions %>%
  mutate(overall = colSums(table_proportions[,2])) %>%
  mutate(proportion = total/overall) %>%
  select(Strata, total, proportion) %>%
  mutate(Participated = 1)
#nonpart
table_proportions2 <- table_plot %>%
  filter(Participated == 0) %>%
  group_by(Strata) %>%
  dplyr::summarise(total = n())
#change to numeric
table_proportions2$total <- as.numeric(table_proportions2$total)
#add a new column with total
table_proportions2 <- table_proportions2 %>%
  mutate(overall = colSums(table_proportions2[,2])) %>%
  mutate(proportion = total/overall) %>%
  select(Strata, total, proportion) %>%
  mutate(Participated = 0)
#put together
plot3data <- bind_rows(table_proportions2, table_proportions) %>%
  mutate(partic = case_when(Participated == 1 ~"TIP Participants",
                            Participated == 0 ~ "non TIP Participants"))
plot3data$'Years of Experience' <- plot3data$Strata


plot3 <- ggplot(plot3data, aes(x= Strata, y = `proportion`, fill = Strata)) +
  geom_bar(stat = "identity") +
  facet_wrap(~partic) +
  labs(title = "Proportion of Strata by Participation") +
  labs(fill = "Years of Experience")+
  ylab("Proportion") +
  theme(axis.text.x.bottom = element_blank()) +
  geom_text(size = 3.5,aes(label = percent(`proportion`), vjust = 1.7)) +
  ylim(0,1)

plot3
```


```{r chunk3.7, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE}


#create proportion tables
table_proportions <- table_plot %>%
  group_by(Strata, Retained, Participated) %>%
  dplyr::summarise(total = n()) %>%
  spread(key = Retained, value = total) %>%
  mutate(totals = `0` + `1`) %>%
  mutate(liklihood_not_retained = `0` / totals) %>%
  mutate(liklihood_retained = `1` / totals) %>%
           mutate(partic = case_when(Participated == 1 ~"TIP Participants",
                            Participated == 0 ~ "non TIP Participants"))
library(scales)

table_proportions$new <- percent(table_proportions$liklihood_retained)

plot2 <- ggplot(table_proportions, aes(x= Strata, y = as.numeric(`liklihood_retained`), fill = Strata)) +
  geom_bar(stat = "identity") +
  facet_wrap(~partic) +
  labs(title = "Liklihood of Retention by Participation and Strata") +
  ylab("Likelihood of Retention") +
  theme(axis.text.x.bottom=element_blank()) +
  geom_text(aes(label = `new`), vjust = 1.6) +
  ylim(0,1)

plot2


```



```{r chunk3.8.2, eval=FALSE, fig.show='hold', message=FALSE, warning=FALSE, include=FALSE}
#plot the other one too
#other way
rm(list = ls()) #clear the environment

##remove the year variables out of each
load("data/merged_participation.Rdata")

#first, create a new yeears_service table with variable

#THOSE WHO DID PARTICIPATE

table_1 <- merged_participation
table_1$`Years of Service` <- as.numeric(table_1$`Years of Service`)
table_1$`Participated` <- as.numeric(table_1$`Participated`)
table_1$Retained <- as.numeric(table_1$Retained)

table_try <- table_1 %>%
  select(`Years of Service`, Retained, Participated) %>%
  filter(Participated == 1) %>%
  group_by(`Years of Service`, Retained, Participated) %>%
  summarise(total_numer = n())
#new data framee with total sums
table_1_totals <- table_1 %>%
  select(`Years of Service`, Retained, Participated) %>%
  filter(Participated == 1) %>%
  group_by(`Years of Service`) %>%
  summarise(total = n ())
#merge that back with the previous
table_try <- merge(table_try, table_1_totals)
#can now do this one of two ways, first by the proportion that was retained, or by the other way
#later do this by parsing out the two
table_try_1 <- table_try %>% 
  mutate(prop = total_numer/total)

table_part <- table_try_1

#THOSE WHO DID NOT PARTICIPATE

#first, create a new yeears_service table with variable
table_1 <- merged_participation
table_1$`Years of Service` <- as.numeric(table_1$`Years of Service`)
table_1$`Participated` <- as.numeric(table_1$`Participated`)
table_1$Retained <- as.numeric(table_1$Retained)

table_try <- table_1 %>%
  select(`Years of Service`, Retained, Participated) %>%
  filter(Participated == 0) %>%
  group_by(`Years of Service`, Retained, Participated) %>%
  summarise(total_numer = n())
#new data framee with total sums
table_1_totals <- table_1 %>%
  select(`Years of Service`, Retained, Participated) %>%
  filter(Participated == 0) %>%
  group_by(`Years of Service`) %>%
  summarise(total = n ())
#merge that back with the previous

table_try <- merge(table_try, table_1_totals)

#can now do this one of two ways, first by the proportion that was retained, or by the other way
#later do this by parsing out the two
table_try_1 <- table_try %>% 
  mutate(prop = total_numer/total)
table_nonpart <- table_try_1


############what is we jsut merged the two tables

tables <- rbind(table_part, table_nonpart)
tables$Participated[tables$Participated == 1] <- "TIP Participants"
tables$Participated[tables$Participated == 0] <- "non-TIP Participants"

plot <- ggplot(data = tables, aes(x=`Years of Service`, y = `prop`)) +
  geom_point(aes(color = as.factor(`Retained`))) +
  facet_wrap(~`Participated`) +
  xlab("Years of Service") +
  ylab("Proportion Retained") +
  labs(color = "Retained") +
  ggtitle("Proportion Retention Breakdown") +
  xlim(0,30)
  
plot


```



## Constructing a dataset for analysis

In order to continue with a statistical analysis, we create an analysis dataset with a row for each Deidentified ID that has all of its unique variables. Recall that the merged_dataset has multiple rows for the same Deidentified ID. For instance, teacher ID #72 may have changed schools over the time we have the data. Or, teacher ID #72 may have changed their title (example: a Special Education teacher moving to a General Education teacher). In order to capture that change, we can come up with a new variable that records the number of times the teacher changed schools, titles, etc.

Variables that change for each ID, such as gender, or school, will be gathered and then re-coded. For instance, the original data set had a unique entry for each year in the data. Here, that is recoded as "total years in TIP data set". Or, some teachers changed gender over the time frame. This is recoded as "number of genders". Or, some teachers changed schools during the time frame. This is recoded as "number of schools".

These new variables, named "number of titles", or "numer of schools", are then merged back with merged_participation, to give us one complete dataset for analysis. Limitations to this approach are discussed later.

We end up with a data set of total of 16012 unique teachers, 312 of whom participanted and 15700 of whom did not participate. An additional note is that the analysis data set here includes a total of 16003 observations, 9 were removed because the variable for Gender was missing.

In choosing the variables to analyze from the analysis dataset, we must make a choice about which variables to choose. For instance, some teachers have more than 1 role, or more than 1 school worked at. In this example code, I simply choose the first observation that appears for that teacher. Since the information is organized by employee age, this is the youngest age at which the teacher started. We can later choose to organize this information differently if need be.


```{r chunk4, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE}

rm(list = ls()) #clear the environment

#load in the dataset
load("data/merged_participation.Rdata")

#We create a new variable called switch_schools that is 1 if they switched schools or 0 if they stayed at the same school. We do the  same for ethnicity and title, capturing how many times teachers changed these variables

#repeats of schools
merged_variables_schools_switch <- merged_participation %>%
  select(Organization, `Deidentified ID`) %>%
  distinct() %>%
  arrange(`Deidentified ID`)
tally_switched_schools <- merged_variables_schools_switch %>%
  group_by(`Deidentified ID`) %>%
  tally() %>%
  mutate("number_of_schools_worked" = n) %>%
  select(`Deidentified ID`, number_of_schools_worked)

#repeats of ethnicity
merged_variables_ethnicity_switch <- merged_participation %>%
  select(Ethnicity, `Deidentified ID`) %>%
  distinct() %>%
  arrange(`Deidentified ID`)
tally_ethnicity_schools <- merged_variables_ethnicity_switch %>%
  group_by(`Deidentified ID`) %>%
  tally() %>%
  mutate("number_of_ethnicities" = n) %>%
  select(`Deidentified ID`, number_of_ethnicities)

#repeats of gender
merged_variables_gender_switch <- merged_participation %>%
  select(Gender, `Deidentified ID`) %>%
  distinct() %>%
  arrange(`Deidentified ID`)
tally_gender_schools <- merged_variables_gender_switch %>%
  group_by(`Deidentified ID`) %>%
  tally() %>%
  mutate("number_of_genders" = n) %>%
  select(`Deidentified ID`, number_of_genders)

#repeats of titles
merged_variables_titles_switch <- merged_participation %>%
  select(Title, `Deidentified ID`) %>%
  distinct() %>%
  arrange(`Title`)
tally_title_schools <- merged_variables_titles_switch %>%
  group_by(`Deidentified ID`) %>%
  tally() %>%
  mutate("number_of_titles" = n) %>%
  select(`Deidentified ID`, number_of_titles)

merged_participation <- merge(merged_participation, tally_switched_schools, by = "Deidentified ID")
merged_participation <- merge(merged_participation, tally_gender_schools, by = "Deidentified ID")
merged_participation <- merge(merged_participation, tally_ethnicity_schools, by = "Deidentified ID")
merged_participation <- merge(merged_participation, tally_title_schools, by = "Deidentified ID")

#save dataset
save(merged_participation, file = "data/merged_participation_analysis.Rdata")

#from the merged_participation_analysis, we can move forward with regressions and analyses in the next chunks of code, by picking out the variables of interest and choosing "distinct"



```



## Model 1


$$log(r/(1-r)) = \beta_0 + \beta_1{x_1} + \beta_2{x_2} +... \beta_p{x_p}$$
The logistic regression model (above) forms the basis for the following statistical tests. Where r is proability of a teacher being retained or not retained, and each x represents a possible predictor, such as gender or years of service in the district, then the model tests the significance of each regression coefficient in predicting the log odds of retention.

First, we test the model using participation in TIP as the predictor and retention as the outcome variable.


$$log(r/(1-r)) = \beta_0 + \beta_1{x_1}$$

$$\beta_1 = Participation $$

The output of Model 1 shows that participation in TIP increases the log odds of retention by 0.35. In other words, TIP teachers are 35% more likely to be retained than non-TIP teachers. This value is statistically significant, given that p <.05.


```{r chunk5, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE}

#use the dataset
rm(list = ls()) #clear the environment
load("data/merged_participation_analysis.Rdata")

#Create a new data frame that chooses teacher status based on the first one in that data set

analysis_data <- merged_participation %>%
  arrange(`Deidentified ID`, `Employee Age`) %>%
  group_by(`Deidentified ID`,Strata, Retained, Participated) %>%
  group_by(`Deidentified ID`) %>%
  filter(row_number() == 1) #choosing the first row for each Deidentified ID

#recode the levels of strata
analysis_data$Strata <- as.factor(analysis_data$Strata)
analysis_data$Strata <- recode_factor(analysis_data$Strata, `1` = "less than 5", `2` = "5 to less than 10 years", `3` = "10 to 15 years", `4` = "greater than or equal to 15")
#make the other variables factors
analysis_data$Gender <- as.factor(analysis_data$Gender)
analysis_data$Ethnicity <- as.factor(analysis_data$Ethnicity)
analysis_data$`Employee Age` <- as.numeric(analysis_data$`Employee Age`)
analysis_data$Retained <- as.numeric(analysis_data$Retained)
analysis_data$Participated <- as.factor(analysis_data$Participated)
analysis_data$Title <- as.factor(analysis_data$Title)

library(dplyr)
#for analysis purposes, we next want to find out where missing data are for participants
#sum(is.na(analysis_data$Gender)) #there are 9 missing values for Gender, so we remove these
analysis_data <- analysis_data[!is.na(analysis_data$Gender),]

#model1: predicting retention by participation in TIP

logit1 <- glm(Retained ~ Participated, data = analysis_data, family = binomial)
summary(logit1)

library(pander)
library(xtable)
table <- logit1$coefficients
kable(table)

```

## Model 2

Next, we test the model controlling for the effects of Gender and Ethnicity. Controlling for ethnicity and participation in TIP, Gender is not a significant predictor of retention. We compared this full model (with the three predictors) against models that only included Gender and Ethnicity, or Participation and Gender, or Participation and Ethnicity, or one of each of the variables. The results from the F test comparing these models indicate that the model with Participation and Gender explains the most variation in Retention Rate, rather than the model with all three predictor variables.

Ethnicity, on the other hand is a significant predictor for 4 of the 8 listed ethnicities. GO ON HERE 



$$log(r/(1-r)) = \beta_0 + \beta_1{x_1} + \beta_2{x_2} + \beta_3{x_3}$$


$$log(r/(1-r)) = \beta_0 + \beta_1{x_1}$$

$$\beta_1 = Participation $$
$$\beta_2 = Gender $$
$$\beta_3 = Ethnicity $$

```{r chunk5.54, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE}

logit2 <- glm(Retained ~ Participated  +Ethnicity , data = analysis_data, family = binomial)
logit2
summary(logit2)
table
kable(table)
```


```{r chunk5.5, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

#determine if Gender is equal to 0, and if the model is better without that predictor
logit2 <- glm(Retained ~ Participated + Ethnicity, data = analysis_data, family = binomial)
fstat4 <- ((deviance(logit2)-deviance(logit1))/(df.residual(logit2)-df.residual(logit1)))/(deviance(logit1)/df.residual(logit1))
1-pf(fstat4,(df.residual(logit2)-df.residual(logit1)),df.residual(logit1)) #not significant

#determine if Ethnicity is equal to 0, and if the model is better without that predictor
logit2 <- glm(Retained ~ Participated + Gender, data = analysis_data, family = binomial)
fstat4 <- ((deviance(logit2)-deviance(logit1))/(df.residual(logit2)-df.residual(logit1)))/(deviance(logit1)/df.residual(logit1))
1-pf(fstat4,(df.residual(logit2)-df.residual(logit1)),df.residual(logit1)) #is significant, so keep Ethnicity

#in this case, I think we should exclude Gender from the model

#run the step function
step(logit1)

```


```{r chunk5.6, eval=FALSE, fig.show='hold', message=FALSE, warning=FALSE, include=FALSE}


#interaction effects

logit4 <- glm(Retained ~ Participated * Gender * Ethnicity, data = analysis_data, family = binomial)
summary(logit4)



```

# Model 3

Next, we examine the interaction effects of Strata and Participation. While strata itself was a signficant predictor of retention, the interaction between strata and participation in TIP was not significant. 

$$log(r/(1-r)) = \beta_0 + \beta_1{x_1} * \beta_2{x_2} $$

$$\beta_1 = Strata $$
$$\beta_2 = Participation $$


```{r chunk12.2, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE}

#model 3: Strata and TIP

logit3 <- glm(Retained ~ Strata * Participated, data = analysis_data, family = binomial)
summary(logit3)
xtable(logit3)

#
logit5 <- glm(Retained ~ Strata * Participated + Ethnicity + Gender, data = analysis_data, family = binomial)
summary(logit5)
xtable(logit5)


```




```{r chunk12, eval=FALSE, fig.show='hold', message=FALSE, warning=FALSE, include=FALSE}
# 
# # Model4:
# 
# Next, we dive deeper into TIP participants specifically. Among TIP participants, does strata (number of prior years in the classroom) affect the impact of TIP on retention rates?
# 
# In order to analyze this question, we subset out only the teachers who participated in TIP, and predict their retention based on Strata levels. Being in Strata 1 (teaching less than 5 years) increases odds of retention by 0.888 (p<.05). Moving from Strata 1 to Strata 2 increases the odds of retention by .76, and this effect is also significant. Moving from the second to the third strata is not a significant predictor of retention. However, being in the final strata decreases chances of retention by odds of -1.08 (p<.05) 
# 
# Then, we look at the individual predictors of each Year of Service on the outcome of retention, only among TIP Participants.



#create new dataframe

analysis_data_participants <- analysis_data %>%
  filter(Participated == 1)

logit3 <- glm(Retained ~ (Strata), data = analysis_data_participants, family = binomial)
summary(logit3)

logit3.5 <- glm(Retained ~ as.factor(`Years of Service`), data = analysis_data_participants, family = binomial)
summary(logit3.5)
```



```{r chunk13, eval=FALSE, fig.show='hold', message=FALSE, warning=FALSE, include=FALSE}

# # Model 5
# 
# Finally, we analyze the effect of other variables, such as gender and ethnicity on retention of TIP participants. None are significant.

#model 

logit4 <- glm(Retained ~ Ethnicity + Gender, data = analysis_data_participants)
summary(logit4)

```

# Discussion/Limitations

The first question of interest is the retention of teachers in SDP (the School District of Philadelphia) who participated or did not particpate in TIP. The analyses show that TIP is a significant predictor of retention for teachers in SDP.

Model 1 shows us that participating in TIP did have a positive effect on teacher retention. 

Model 4 demonstrates how this effect varied for different strata and year levels. This resonates with the visual nature of the plots and graphs, which show that the liklihood of retention for TIP teachers changes in each Strata. 

In model 3.5, we also analyze the predictor variables on years of service as predictors of teacher retention. Teachers who served for 1,3, and 4 years were more likely to leave than teachers who stayed for 1 years, as were teachers who stayed for 13,21, and 22 years. 

Next steps will be to look more closely at the differences in teacher retention rates betweeen strata and number of years served for TIP teachers. As mentioned in the original proposal, "non-retention of high-performing teachers is a serious problem for urban school districts". Better understanding the patterns of retention based on strata and years worked can shed light the specific groups that a program like TIP has the most effect on, and better yet, why, such a program impacts retention rates.

A few limitations on the current analysis exist. First, the sample size of TIP teachers is a fraction of the sample size of non_TIP teachers. Such sample size differences may result in complications for future interaction analyses or causal interpretations. Additionally, teacher retention is calculated as a teacher staying in the school district dataset from one year to another. This fails to capture differencees in changed Titles of teachers, or where specifically teachers move throughout the system.

Lastly, an additional variable that would help to answer the original research questions posed in the proposal would be an indicator of the date that teachers began the TIP program. As it stands, the only information that we have from TIP is the Deidentified IDs of participants, which were matched with administrative data from the 8 year period. If given an indicator of the start date and/or end date that teachers participated in TIP, we could analyze trends like amount of time participated in TIP, and it's effect on retention.